\begin{thebibliography}{99}
\small{

% % % % % % %
% % % % % % %\bibitem{AADGMS93} 
% % % % % % %Afek Y., Attiya H., Dolev D., Gafni E., Merritt M. and Shavit N., 
% % % % % % %Atomic Snapshots of Shared Memory. 
% % % % % % %{\em Journal of the ACM}, 40(4):873-890, 1993.



\bibitem{1133985}
% Ali-Reza Adl-Tabatabai, Brian~T. Lewis, Vijay Menon, Brian~R. Murphy, Bratin
%   Saha, and Tatiana Shpeisman.
Adl-Tabatabai A.-R., Lewis B.~T., Menon V., Murphy B.~R., 
  Saha B. and Shpeisman T.
\newblock Compiler and runtime support for efficient software transactional
  memory.
\newblock In {\em PLDI '06: Proceedings of the 2006 ACM SIGPLAN conference on
  Programming language design and implementation}, pages 26--37, New York, NY,
  USA, 2006. ACM.



\bibitem{AVL62}
Adelson-Velskii G. and Landis E.~M.
\newblock An algorithm for the organization of information.
\newblock In {\em Proc. of the USSR Academy of Sciences}, volume 146, pages
  263--266, 1962.






\bibitem{afek10}
 Afek, Y.,  Avni, H.,   Dice, D. and  Shavit, N.: Efficient lock free privatization. 
In: {\it Proc.  14th Int'l conference on Principles of Distributed Systems 
(OPODIS'10)}, pp. 333--347, Springer-Verlag,  LNCS \#6490 (2010) 



%\vspace{-0.2cm}
\bibitem{ADT95}
Afek Y., Dauber D. and Touitou D.,
Wai-free made Fast. 
{\it Proc. 7th  Int'l  ACM Symposium on Parallelism in Algorithms 
and Architectures   (SPAA'10)},  ACM Press, pp. 538-547,  1995. 






\bibitem{ALS09}
%Kunal Agrawal, I-Ting~Angelina Lee, and Jim Sukha.
Agrawal K., Lee I-T.~A. and Sukha J.
\newblock Safe open-nested transactions through ownership.
\newblock In {\em Proc. of the 14th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2009.


% \bibitem{1378553}
% Kunal Agrawal, I-Ting~Angelina Lee, and Jim Sukha.
% \newblock Safe open-nested transactions through ownership.
% \newblock In {\em SPAA '08: Proceedings of the twentieth annual symposium on
%   Parallelism in algorithms and architectures}, pages 110--112, New York, NY,
%   USA, 2008. ACM.



\bibitem{Alvisi_lock-freeserializable}
Alvisi L.
\newblock Lock-free serializable transactions.
\newblock Technical report, 2005.



%\vspace{-0.2cm}
\bibitem{AM99}
Anderson J. and Moir M., 
Universal Constructions for Large Objects. 
{\it  IEEE Transactions on Parallel and Distributed  Systems},
10(12):1317-1332,  1999.



%\vspace{-0.2cm}
\bibitem{ALKJKW09}
Ansar M.,  Luj\'an M.,  Kotselidis Ch., Jarvis K.,  Kirkham Ch. 
and  Watson Y., 
Steal-on-abort: Dynamic Transaction Reordering to
Reduce Conflicts in Transactional Memory. 
{\it 4th  Int'l ACM Sigplan Conference on  High Performance Embedded  
Architectures and Compilers (HiPEAC'09)}, ACM Press,   pp. 4-18, 2009.


% \bibitem{1505821}
% Mohammad Ansari, Mikel Luj\'{a}n, Christos Kotselidis, Kim Jarvis, Chris
%   Kirkham, and Ian Watson.
% \newblock Steal-on-abort: Improving transactional memory performance through
%   dynamic transaction reordering.
% \newblock In {\em HiPEAC '09: Proceedings of the 4th International Conference
%   on High Performance Embedded Architectures and Compilers}, pages 4--18,
%   Berlin, Heidelberg, 2009. Springer-Verlag.




\bibitem{1146428}
% Hagit Attiya, Leah Epstein, Hadas Shachnai, and Tami Tamir.
Attiya H., Epstein L., Shachnai H. and Tamir T.
\newblock Transactional contention management as a non-clairvoyant scheduling
  problem.
\newblock In {\em PODC '06: Proceedings of the twenty-fifth annual ACM
  symposium on Principles of distributed computing}, pages 308--315, New York,
  NY, USA, 2006. ACM.

\bibitem{1538908}
% Hagit Attiya, Rachid Guerraoui, Danny Hendler, and Petr Kuznetsov.
Attiya H., Guerraoui R., HendlerD. and Kuznetsov P.
\newblock The complexity of obstruction-free implementations.
\newblock {\em J. ACM}, 56(4):1--33, 2009.



\bibitem{AH11}
Attiya  H. and Hillel E.,
Single-version STM Can be Multi-version Permissive. 
{\it Proc. 12th Int'l Conference on Distributed Computing and Networking
(ICDCN'11)}, Springer-Verlag, LNCS \#6522, pp.~83-94, 2011.


\bibitem{1584015}
Attiya H., Hillel E. and Milani A.
\newblock Inherent limitations on disjoint-access parallel implementations of
  transactional memory.
\newblock In {\em SPAA '09: Proceedings of the twenty-first annual symposium on
  Parallelism in algorithms and architectures}, pages 69--78, New York, NY,
  USA, 2009. ACM.




%\vspace{-0.2cm}
\bibitem{AM09}
Attiya H. and Milani A., 
Transactional Scheduling for Read-Dominated Workloads. 
{\it 13th Int'l Conference on  Principles of Distributed Systems
(OPODIS'09)},  Springer Verlag  LNCS \#5923, pp. 3-17, 2009. 


% \bibitem{1696831}
% Hagit Attiya and Alessia Milani.
% \newblock Transactional scheduling for read-dominated workloads.
% \newblock In {\em OPODIS '09: Proceedings of the 13th International Conference
%   on Principles of Distributed Systems}, pages 3--17, Berlin, Heidelberg, 2009.
%   Springer-Verlag.







\bibitem{BM93}
Babao\u{g}lu \"{O}. and Marzullo K., 
Consistent Global States of Distributed Systems: Fundamental
Concepts and Mechanisms. Chapter~4 in ``Distributed Systems''.  
ACM Press, Frontier Series,  pp~55-93, 1993.


\bibitem{Bal06}
% Lucia Ballard.
Ballard L.
\newblock Conflict avoidance: Data structures in transactional memory, May
  2006.
\newblock Undergraduate thesis, Brown University.

% \bibitem{Bal06}
% L.~Ballard.
% \newblock Conflict avoidance: Data structures in transactional memory, May
%   2006.
% \newblock Undergraduate thesis, Brown University.


\bibitem{Bay72}
% Rudolf Bayer.
Bayer R.
\newblock Symmetric binary b-trees: Data structure and maintenance algorithms.
\newblock {\em Acta Informatica 1}, 1(4):290--306, 1972.



\bibitem{BSW79}
Bernstein Ph.A., Shipman D.W. and Wong W.S., 
Formal Aspects of Serializability in Database Concurrency Control. 
{\it IEEE Transactions on Software Engineering}, SE-5(3):203-216, 1979. 


\bibitem{Borkar2007}
Borkar S.
\newblock Thousand core chips: a technology perspective.
\newblock In {\em DAC}, pages 746--749, 2007.


% \bibitem{Borkar2007}
% S.~Borkar.
% \newblock Thousand core chips: a technology perspective.
% \newblock In {\em DAC}, pages 746--749, 2007.



%\vspace{-0.2cm}
\bibitem{BG93}
Borowsky E. and Gafni E., 
Generalized FLP Impossibility Results for $t$-Resilient Asynchronous
Computations.
{\it Proc. 25th ACM  Symposium on Theory of Computing (STOC'93)}, 
ACM Press, pp. 91-100, 1993.



\bibitem{IRISAppr}
% Luc Boug\'e, Joaquim Gabarro, Xavier Messeguer, and Nicolas Schabanel.
Boug\'e L., Gabarro J., Messeguer X. and Schabanel N.
\newblock Height-relaxed {AVL} rebalancing: A unified, fine-grained approach to
  concurrent dictionaries, 1998.
\newblock Research Report 1998-18, ENS Lyon.

% \bibitem{IRISAppr}
% L.~Boug\'e, J.~Gabarro, X.~Messeguer, and N.~Schabanel.
% \newblock Height-relaxed {AVL} rebalancing: A unified, fine-grained approach to
%   concurrent dictionaries.
% \newblock Technical Report RR1998-18, ENS Lyon, 1998.





\bibitem{BCCO10}
%Nathan~G. Bronson, Jared Casper, Hassan Chafi, and Kunle Olukotun.
Bronson N.~G., Casper J., Chafi H. and Olukotun K.
\newblock A practical concurrent binary search tree.
\newblock In {\em Proc. of the 15th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2010.

% \bibitem{BCCO10}
% N.~G. Bronson, J.~Casper, H.~Chafi, and K.~Olukotun.
% \newblock A practical concurrent binary search tree.
% \newblock In {\em Proc. of the 15th ACM SIGPLAN Symp. on Principles and
%   Practice of Parallel Programming}, 2010.



\bibitem{BGK12}
% Victor Bushkov, Rachid Guerraoui, and Michał Kapałka.
Bushkov V., Guerraoui R. and Kapa\l{}ka M.
\newblock On the liveness of transactional memory.
\newblock {\it Proc. 7th  Int'l ACM symposium on Principles of distributed computing (PODC '12)}.
 ACM, New York, USA, Pages 9-18 2012.



\bibitem{CR06}
Cachopo J. and Rito-Silva A., 
Versioned Boxes as the Basis for Transactional Memory. 
{\it Science of Computer Progr.}, 63(2):172-175, 2006. 

% \bibitem{1228566}
% {a}o Cachopo, Jo\ and Ant\'{o}nio Rito-Silva.
% \newblock Versioned boxes as the basis for memory transactions.
% \newblock {\em Sci. Comput. Program.}, 63(2):172--185, 2006.




\bibitem{CCKO08}
%Chi Cao~Minh, JaeWoong Chung, Christos Kozyrakis, and Kunle Olukotun.
Cao~Minh C., Chung J., Kozyrakis C. and Olukotun K.
\newblock {STAMP}: Stanford transactional applications for multi-processing.
\newblock In {\em Proc. of The IEEE Int'l Symp. on Workload Characterization},
  2008.



%\vspace{-0.2cm}
\bibitem{CER10}
Chuong Ph., Ellen F. and Ramachandran V.,
A Universal Construction for Wait-free Transaction Friendly Data Structures.
{\it Proc. 22th  Int'l  ACM Symposium on Parallelism in Algorithms 
and Architectures   (SPAA'10)},  ACM Press, pp. 335-344,  2010. 

\bibitem{CH05}
%Christopher Cole and Maurice Herlihy.
Cole C. and Herlihy M.~P.
\newblock Snapshots and software transactional memory.
\newblock {\em Sci. Comput. Program.}, 58(3):310--324, 2005.



\bibitem{CGR12}
Crain T., Gramoli V. and Raynal M.
\newblock A speculation-friendly binary search tree.
\newblock In {\em Proc. of the 17th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2012.

% \bibitem{CGR12}
% T.~Crain, V.~Gramoli, and M.~Raynal.
% \newblock A speculation-friendly binary search tree.
% \newblock In {\em PPoPP}, 2012.



\bibitem{CIR10}
Crain T., Imbs D. and Raynal M.,
Read invisibility, virtual world consistency and  permissiveness
are compatible.
{\it Tech Report \#1958}, IRISA, Univ. de  Rennes 1, France, November 2010. 



\bibitem{CIR11}
Crain T., Imbs D. and Raynal M.,
Towards a universal construction for transaction-based multiprocess programs.
{\it Proceedings of the 13th International Conference on Distributed Computing and Networking (ICDCN 2012)}, 
Springer-Verlag LNCS, 2012.

% \bibitem{CIR12} 
% Crain, T., Imbs, D.,  Raynal, M.: 
% Towards a Universal Construction for Transaction-based Multiprocess Programs.
% In: {\it Proc. 13th Int'l Conference on Distributed Computing and Networking (ICDCN'12)}, 
% pp. 61--75, Springer Verlag LNCS \#7129 (2012) 



\bibitem{DSS10}
Dalessandro L., Spear M. and Scott M.~L.
\newblock {NO}rec: streamlining {STM} by abolishing ownership records.
\newblock In {\em Proc. of the 15th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2010.





\bibitem{DS09}
Dalessandro, L. and Scott, M.:
Strong Isolation is a Weak Idea. 
In: {\it Proc. Workshop on transactional memory (TRANSACT'09)} (2009)




\bibitem{DMM+01}
Detlefs D.~L., Martin P.~A., Moir M. and Steele Jr G.~L.
\newblock Lock-free reference counting.
\newblock In {\em PODC}, pages 190--199, 2001.


\bibitem{dice10}
Dice, D., Matveev, A. and  Shavit, N.:
Implicit privatization using private transactions. 
In: {\it Proc. Workshop on transactional memory (TRANSACT'10)} (2010)


\bibitem{DSS06}
Dice D., Shalev O. and Shavit N.,
Transactional Locking II.
{\em Proc. 20th Int'l Symposium on Distributed Computing (DISC'06)},
Springer-Verlag, LNCS \#4167, pp.~194-208, 2006.

% \bibitem{DSS06}
% D.~Dice, O.~Shalev, , and N.~Shavit.
% \newblock Transactional locking {II}.
% \newblock In {\em Proc. of the 20th Int'l Symp. on Distributed Computing},
%   2006.
% 
% \bibitem{DSS06}
% D.~Dice, O.~Shalev, , and N.~Shavit.
% \newblock Transactional locking {II}.
% \newblock In {\em Proc. of the 20th Int'l Symp. on Distributed Computing},
%   2006.
% 
% \bibitem{dice06}
% Dice, D., Shalev O., Shavit, N.:
% Transactional Locking II.
% In: {\em Proc. 20th Int'l Symp. on Distributed Computing (DISC'06)},
% Springer-Verlag, LNCS \#4167, pp.~194--208 (2006)
% 
% 
% \bibitem{Dice06transactionallocking}
% Dave Dice, Ori Shalev, and Nir Shavit.
% \newblock Transactional locking ii.
% \newblock In {\em In Proc. of the 20th Intl. Symp. on Distributed Computing},
%   2006.




%\vspace{-0.2cm}
\bibitem{D68}
Dijkstra E.~W., 
Solution of a Problem in Concurrent Programming Control.
{\it  Communications of the  ACM}, 8(9):69, 1968.

\bibitem{DLM78}
Dijkstra E.~W., Lamport L., Martin A.~J., Scholten C.~S. and 
  Steffens E.~F.~M.
\newblock On-the-fly garbage collection: an exercise in cooperation.
\newblock {\em Commun. ACM}, 21(11):966--975, 1978.

% \bibitem{DLM78}
% Edsger~W. Dijkstra, Leslie Lamport, A.~J. Martin, C.~S. Scholten, and E.~F.~M.
%   Steffens.
% \newblock On-the-fly garbage collection: an exercise in cooperation.
% \newblock {\em Commun. ACM}, 21(11):966--975, 1978.



\bibitem{DFGG11}
% Aleksandar Dragojevic, Pascal Felber, Vincent Gramoli, and Rachid Guerraoui.
Dragojevi\'{c} A., Felber P., Gramoli V. and Guerraoui R.
\newblock Why {STM} can be more than a research toy.
\newblock {\em Commun. ACM}, 54(4):70--77, 2011.

% \bibitem{LPD-REPORT-2009-003}
% Aleksandar Dragojevic, Pascal Felber, Vincent Gramoli, and Rachid Guerraoui.
% \newblock Why {STM} can be more than a {R}esearch {T}oy.
% \newblock Technical report, 2009.



%\vspace{-0.2cm}
\bibitem{DGK09}
Dragojevi\'{c} A., Guerraoui R. and  Kapa\l{}ka M., 
Stretching Transactional Memory. 
{\it  Proc. Int'l  2009 ACM SIGPLAN conference on Programming
 language design and implementation (PLDI '09)},  
ACM Press, pp. 155-165,  2009. 

% \bibitem{1542494}
% Aleksandar Dragojevi\'{c}, Rachid Guerraoui, and Michal Kapalka.
% \newblock Stretching transactional memory.
% \newblock In {\em PLDI '09: Proceedings of the 2009 ACM SIGPLAN conference on
%   Programming language design and implementation}, pages 155--165, New York,
%   NY, USA, 2009. ACM.




\bibitem{DHW97}
Dwork C., Herlihy M.P. and Waarts O.
\newblock Contention in shared memory algorithms.
\newblock {\em J. ACM}, 44:779--805, November 1997.


\bibitem{EFRB10}
Ellen F., Fatourou P., Ruppert E. and Breugel F.~van.
\newblock Non-blocking binary search trees.
\newblock In {\em Proceedings of the 29th ACM SIGACT-SIGOPS symposium on
  Principles of distributed computing}, PODC '10, pages 131--140, New York, NY,
  USA, 2010. ACM.


\bibitem{vEKZ76}
Emde~Boas P.~van, Kaas R., and Zijlstra E.
\newblock Design and implementation of an efficient priority queue.
\newblock {\em Theory of Computing Systems}, 10:99--127, 1976.
\newblock 10.1007/BF01683268.




\bibitem{Ennals05efficientsoftware}
% Robert Ennals and Robert Ennals.
Ennals R.
\newblock Efficient software transactional memory.
\newblock Technical report, 2005.

\bibitem{Ennals06softwaretransactional}
% Robert Ennals and Robert Ennals.
Ennals R.
\newblock Software transactional memory should not be obstruction-free.
\newblock Technical report, 2006.





%\vspace{-0.2cm}
\bibitem{FK09}
Fatourou P. and Kallimanis N., 
The Red-blue Adaptive Universal Construction. 
{\it  Proc. 22nd  Int'l  Symposium on Distributed Computing
(DISC '09)},  Springer-Verlag, LNCS\#5805, pp. 127-141,  2009. 



%\vspace{-0.2cm}
\bibitem{F11}
Felber P., 
Compiler Support for STM Systems. 
Lecture given at the  {\it TRANSFORM  Initial Training School},  University
of Rennes  1 (France), 7-11 February 2011. 






\bibitem{FFGH08}
Felber P., Fetzer Ch., Guerraoui R. and Harris T., 
Transactions are coming Back, but Are They The Same?
{\it ACM Sigact News, DC Column}, 39(1):48-58, 2008.

%\vspace{-0.2cm}
\bibitem{FFR08}
Felber P.,  Fetzer Ch. and  Riegel T., 
Dynamic Performance Tuning of Word-Based Software Transactional Memory. 
{\it  Proc. 13th  Int'l  ACM SIGPLAN Symposium on Principles and
 Practice of Parallel Programming (PPoPP'08) },  ACM Press,pp. 237-246,  2008. 

% \bibitem{FFR08}
% Pascal Felber, Christof Fetzer, and Torvald Riegel.
% \newblock Dynamic performance tuning of word-based software transactional
%   memory.
% \newblock In {\em Proc. of the 13th ACM SIGPLAN Symp. on Principles and
%   Practice of Parallel Programming}, 2008.



\bibitem{FGG09}
% Pascal Felber, Vincent Gramoli, and Rachid Guerraoui.
Felber P., Gramoli V. and Guerraoui R.
\newblock Elastic transactions.
\newblock In {\em Proc. of the 23rd Int'l Symp. on Distributed Computing},
  2009.

% \bibitem{FGG09}
% P.~Felber, V.~Gramoli, and R.~Guerraoui.
% \newblock Elastic transactions.
% \newblock In {\em Proc. of the 23rd Int'l Symp. on Distributed Computing},
%   2009.
% 
% \bibitem{FGG09}
% P.~Felber, V.~Gramoli, and R.~Guerraoui.
% \newblock Elastic transactions.
% \newblock In {\em DISC}, 2009.
% 
% 
% \bibitem{LPD-REPORT-2009-002}
% Pascal Felber, Vincent Gramoli, and Rachid Guerraoui.
% \newblock Elastic {T}ransactions.
% \newblock Technical report, 2009.


% \bibitem{1360456}
% Pascal Felber, Christof Fetzer, Rachid Guerraoui, and Tim Harris.
% \newblock Transactions are back---but are they the same?
% \newblock {\em SIGACT News}, 39(1):48--58, 2008.

% %\vspace{-0.2cm}
% \bibitem{FFGH08}
% Felber P., Fetzer Ch., Guerraoui R. and Harris T., 
% Transactions are Coming Back, but Are They The Same?
% {\it ACM Sigact News, Distributed  Computing Column}, 39(1):48-58, 2008.


% % % % % % % %\bibitem{FH07}  %  presents the WSTM system (not permissive)
% % % % % % % %Fraser T. and Harris T.,
% % % % % % % %Concurrent Programming Without Locks. 
% % % % % % % %{\it ACM Transactions on Computer Systems}, 25(2), Article 5, 2007. 







\bibitem{FR04}
Fomitchev M. and Ruppert E.
\newblock Lock-free linked lists and skip lists.
\newblock In {\em Proceedings of the twenty-third annual ACM symposium on
  Principles of distributed computing}, PODC '04, pages 50--59, New York, NY,
  USA, 2004. ACM.

% \bibitem{FR04}
% M.~Fomitchev and E.~Ruppert.
% \newblock Lock-free linked lists and skip lists.
% \newblock In {\em PODC}, pages 50--59, 2004.

\bibitem{Fra03}
Fraser K.
\newblock {\em Practical lock freedom}.
\newblock PhD thesis, Cambridge University, September 2003.

% \bibitem{Fra03}
% K.~Fraser.
% \newblock {\em Practical lock freedom}.
% \newblock PhD thesis, Cambridge University, September 2003.



\bibitem{Fraser:2007:CPW:1233307.1233309}
% Keir Fraser and Tim Harris.
Fraser K. and Harris T.
\newblock Concurrent programming without locks.
\newblock {\em ACM Trans. Comput. Syst.}, 25, May 2007.


\bibitem{FLPR99}
Frigo M., Leiserson C., Prokop H. and Ramachandran S.
\newblock Cache-oblivious algorithms.
\newblock In {\em Proceedings of the 40th Annual Symposium on Foundations of
  Computer Science}, pages 285 --297, 1999.






\bibitem{FG01}
Fr{\o}lund S. and Guerraoui R.,
X-Ability: a Theory of Replication.
{\it Distributed Computing}, 14(4):231-249, 2001. 




\bibitem{GG11}
% Vincent Gramoli and Rachid Guerraoui.
Gramoli V. and Guerraoui R.
\newblock Democratizing transactional programming.
\newblock In {\em Proc. of the ACM/IFIP/USENIX 12th Int'l Middleware
  Conference}, 2011.

\bibitem{LPD-ARTICLE-2009-004}
% Vincent Gramoli, Derin Harmanci, and Pascal Felber.
Gramoli V., Harmanci D., Felber P.
\newblock On the {I}nput {A}cceptance of {T}ransactional {M}emory.
\newblock {\em Parallel {P}rocessing {L}etters}, 2009.




\bibitem{GC96}
Greenwald M. and Cheriton D.
\newblock The synergy between non-blocking synchronization and operating system structure.
\newblock In {\em Proceedings of the Second Symposium on Operating System
Design and Implementation}, pages 123--136, 1996.





\bibitem{guerraoui05polymorphic/LPD}
Guerraoui R., Herlihy M.P., and Pochon B.
\newblock Polymorphic {C}ontention {M}anagement.
\newblock In {\em Proceedings of the 19th {I}nternational {S}ymposium on
  {D}istributed {C}omputing ({DISC}'05)}, volume 3724 of {\em Lecture Notes in
  Computer Science}, pages 303--323, 2005.



%\vspace{-0.2cm}
\bibitem{GHP05}
Guerraoui R.,  Herlihy M. and  Pochon B., 
 Towards a Theory of Transactional Contention Managers. 
{\it  Proc. 24th  Int'l  ACM Symposium on Principles of  Distributed 
Computing (PODC'05)},  ACM Press, pp. 258-264,  2005. 


% \bibitem{1073863}
% Rachid Guerraoui, Maurice Herlihy, and Bastian Pochon.
% \newblock Toward a theory of transactional contention managers.
% \newblock In {\em PODC '05: Proceedings of the twenty-fourth annual ACM
%   symposium on Principles of distributed computing}, pages 258--264, New York,
%   NY, USA, 2005. ACM.



% % % % % % % % %%\vspace{-0.2cm}
% % % % % % % % %\bibitem{GHKP05}
% % % % % % % % %Guerraoui R.,  Herlihy M., Kapa\l{}ka M. and  Pochon B., 
% % % % % % % % %Robust Contention Management in Software Transactional Memory. 
% % % % % % % % %{\it  Proc.  OOPSLA 2005 Workshop on Synchronization and Concurrency 
% % % % % % % % %in Object-Oriented Languages (SCOOL'05)},  2005. 


%\vspace{-0.2cm}
\bibitem{GKK08}
Guerraoui R., Kapa\l{}ka M. and  Kouznetsov P., 
The Weakest Failure Detectors to Boost Obstruction-freedom. 
{\it Distributed Computing}, 20(6):415-433, 2008. 


%\vspace{-0.2cm}
\bibitem{GK10}
Guerraoui R. and  Kapa\l{}ka M.,
Principles of Transactional Memory. 
{\it Synthesis Lectures on Distributed Computing Theory},
 Morgan \& Claypool Publishers, 180 pages, 2010. 




\bibitem{GHS08}
Guerraoui R., Henzinger T.A., Singh V.,
Permissiveness  in Transactional Memories. 
{\em Proc. 22th Int'l Symposium on Distributed Computing (DISC'08)},
Springer-Verlag, LNCS \#5218, pp.~305-318, 2008.


% %\vspace{-0.2cm}
% \bibitem{GHS08}
% Guerraoui R.,  Henzinger Th. A. and Singh V., 
% Permissiveness in Transactional Memories. 
% {\it  Proc. 22nd  Int'l  Symposium on Distributed Computing
% (DISC '08) },  Springer-Verlag, LNCS\#5218, pp. 305-319,  2008. 
% 
% 
% \bibitem{LPD-CONF-2008-031}
% Rachid Guerraoui, Thomas~A. Henzinger, and Vasu Singh.
% \newblock Permissiveness in {T}ransactional {M}emories.
% \newblock In {\em Proceedings of the 22nd {I}nternational {S}ymposium on
%   {D}istributed {C}omputing}, 2008.




\bibitem{GK08}
Guerraoui R. and Kapa\l{}ka M., 
On the Correctness of Transactional Memory. 
{\it Proc. 13th ACM SIGPLAN Symposium on Principles and Practice of 
Parallel Programming (PPoPP'08)}, ACM Press, pp. 175-184, 2008. 

% \bibitem{LPD-CONF-2007-017}
% Rachid Guerraoui and Michal Kapalka.
% \newblock On the {C}orrectness of {T}ransactional {M}emory.
% \newblock In {\em Proceedings of the 13th {ACM} {SIGPLAN} {S}ymposium on
%   {P}rinciples and {P}ractice of {P}arallel {P}rogramming ({PP}o{PP}'08)},
%   2008.



% \bibitem{guerraoui08}
% Guerraoui, R., Kapalka, M.:  On the correctness of transactional memory. 
% In: {\it  Proc. 13th ACM SIGPLAN Symposium on Principles and Practice of 
% Parallel Programming (PPoPP '08)},  ACM Press, pp.  175--184 (2008)


\bibitem{GS78}
Guibas L.~J.  and Sedgewick R.
\newblock A dichromatic framework for balanced trees.
\newblock In {\em Proc. of the 19th Annual Symp. on Foundations of Computer
  Science}, 1978.

% \bibitem{GS78}
% L.~J. Guibas and R.~Sedgewick.
% \newblock A dichromatic framework for balanced trees.
% \newblock In {\em Proc. of the 19th Annual Symp. on Foundations of Computer
%   Science}, 1978.

\bibitem{Har01}
Harris T.
\newblock A pragmatic implementation of non-blocking linked-lists.
\newblock In {\em DISC}, pages 300--314, 2001.


% \bibitem{Har01}
% T.~Harris.
% \newblock A pragmatic implementation of non-blocking linked-lists.
% \newblock In {\em DISC}, pages 300--314, 2001.




\bibitem{HCUAGSV07}
Harris T., Cristal  A., Unsal O.S., Ayguade E., Gagliardi  F., Smith B. and
Valero M., 
Transactional Memory: an Overview. 
{\it IEEE Micro}, 27(3):8-29, 2007. 


% %\vspace{-0.2cm}
% \bibitem{H07}
% Harris T., Cristal  A., Unsal O.S., Ayguade E., Gagliardi  F., Smith B. and
% Valero M., 
% Transactional Memory: an Overview. 
% {\it IEEE Micro}, 27(3):8-29, 2007. 
% 
% 
% \bibitem{10.1109/MM.2007.63}
% Tim Harris, Adrian Cristal, Osman~S. Unsal, Eduard Ayguade, Fabrizio Gagliardi,
%   Burton Smith, and Mateo Valero.
% \newblock Transactional memory: An overview.
% \newblock {\em IEEE Micro}, 27:8--29, 2007.



\bibitem{harris06}
 Harris, T.,  Larus, J. and  Rajwar, R.:
Transactional Memory, 
{\it 2nd edition, Synthesis Lectures on Computer Architecture},
Morgan \& Claypool Publishers (2006)



\bibitem{HMPH05}
% Tim Harris, Simon Marlow, Simon Peyton-Jones, and Maurice Herlihy.
Harris T., Marlow S., Peyton-J. S. and Herlihy M.P.
\newblock Composable memory transactions.
\newblock In {\em Proc. of the 10th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2005.




%\vspace{-0.2cm}
\bibitem{H91}
Herlihy M.P., 
Wait-Free Synchronization. 
{\it ACM Transactions on Programming Languages and Systems}, 
13(1):124-149, 1991. 



\bibitem{HK08}
%Maurice Herlihy and Eric Koskinen.
Herlihy M.~P. and Koskinen E.
\newblock Transactional boosting: A methodology for highly-concurrent
  transactional objects.
\newblock In {\em Proc. of the 13th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2008.

% \bibitem{HK08}
% M.~Herlihy and E.~Koskinen.
% \newblock Transactional boosting: A methodology for highly-concurrent
%   transactional objects.
% \newblock In {\em Proc. of the 13th ACM SIGPLAN Symp. on Principles and
%   Practice of Par. Prog.}, 2008.




\bibitem{HLL07}
Herlihy M.P., Lev Y., Luchangco V. and Shavit N.
\newblock A simple optimistic skiplist algorithm.
\newblock In {\em Proceedings of the 14th international conference on
  Structural information and communication complexity}, SIROCCO'07, pages
  124--138, Berlin, Heidelberg, 2007. Springer-Verlag.



\bibitem{HL08}
Herlihy M.P.  and Luchangco V.,
Distributed Computing and the Multicore Revolution.
{\it ACM SIGACT News, DC Column}, 39(1): 62-72, 2008.


% %\vspace{-0.2cm}
% \bibitem{HL08}
% Herlihy M.P.  and Luchangco V.,
% Distributed Computing and the Multicore Revolution.
% {\it ACM SIGACT News}, 39(1): 62-72, 2008.



%\vspace{-0.2cm}
\bibitem{HLMS03}
Herlihy M.,  Luchangco V.,  Moir M. and  Scherer III W.M., 
 Software Transactional Memory for Dynamic-Sized Data Structures. 
{\it  Proc. 22nd  Int'l  ACM Symposium on Principles of  Distributed 
Computing (PODC'03)},  ACM Press, pp. 92-101,  2003. 

% \bibitem{872048}
% Maurice Herlihy, Victor Luchangco, Mark Moir, and William~N. Scherer, III.
% \newblock Software transactional memory for dynamic-sized data structures.
% \newblock In {\em PODC '03: Proceedings of the twenty-second annual symposium
%   on Principles of distributed computing}, pages 92--101, New York, NY, USA,
%   2003. ACM.
% 
% 
% \bibitem{HLMS03}
% Maurice Herlihy, Victor Luchangco, Mark Moir, and William~N. Scherer, III.
% \newblock Software transactional memory for dynamic-sized data structures.
% \newblock In {\em Proc. of the 22nd Annual ACM SIGACT-SIGOPS Symp. on
%   Principles of Distributed Computing}, 2003.


% % % % % % % % % % % %\bibitem{HLMS03}  %  presents the DSTM system (not permissive)
% % % % % % % % % % % %Herlihy M.P.,  Luchangco V., Moir M. and Scherer W.N., 
% % % % % % % % % % % %Software Transactional Memory for Dynamic-size Data Structures. 
% % % % % % % % % % % %{\em Proc. 22th ACM  Symposium on Principles of Distributed Computing 
% % % % % % % % % % % %(PODC'03)},  ACM Press,  pp.~92-101, 2003.


\bibitem{HM93}
Herlihy M.P.  and Moss J.E.B., 
Transactional Memory: Architectural Support for Lock-free Data Structures. 
{\it Proc. 20th ACM Int'l Symposium  on Computer Archictecture (ISCA'93)}, 
pp. 289-300, 1993. 

% \bibitem{HM93}
% M.~Herlihy and J.~E.~B. Moss.
% \newblock Transactional memory: {Architectural} support for lock-free data
%   structures.
% \newblock In {\em Proc. of the 20th Annual Int'l Symp. on Computer
%   Architecture}, 1993.

% %\vspace{-0.2cm}
% \bibitem{HM93}
% Herlihy M.P.  and Moss J.E.B., 
% Transactional Memory: Architectural Support for Lock-free Data Structures. 
% {\it Proc. 20th ACM Int'l Symposium  on Computer Architecture  (ISCA'93)}, 
% ACM Press, pp. 289-300, 1993. 
% 
% 
% \bibitem{herlihy93}
% Herlihy, M., Moss J.M.B.: Transactional memory: architectural support for lock-free data structures, 
% In: {\it Proc.  of the 20th annual Int'l Symposium on Computer Architecture 
% (ISCA '93)}, ACM Press, pp. 289--300 (1993)
% 
% 
% \bibitem{165164}
% Maurice Herlihy and J.~Eliot~B. Moss.
% \newblock Transactional memory: architectural support for lock-free data
%   structures.
% \newblock {\em SIGARCH Comput. Archit. News}, 21(2):289--300, 1993.




%\vspace{-0.2cm}
\bibitem{HS08}
Herlihy M.P.  and Shavit N.,
The Art of Multiprocessor Programming. 
{\it Morgan Kaufmann Pub.}, San Francisco (CA), 508 pages, 2008. 

% \bibitem{HS08}
% M.~Herlihy and N.~Shavit.
% \newblock {\em The Art of Multiprocessor Programming}.
% \newblock Morgan Kauffman, February 2008.



\bibitem{HW90}
Herlihy M.P.  and Wing J.M.,
Linearizability: a Correctness Condition for Concurrent Objects. 
{\it ACM Transactions on Programming Languages and Systems}, 
12(3):463-492, 1990. 

% %\vspace{-0.2cm}
% \bibitem{HW90}
% Herlihy M.P.  and Wing J.M.,
% Linearizability: a Correctness Condition for Concurrent Objects. 
% {\it ACM Transactions on Programming Languages and Systems}, 
% 12(3):463-492, 1990.
% 
% 
% \bibitem{HW90}
% M.~P. Herlihy and J.~M. Wing.
% \newblock Linearizability: a correctness condition for concurrent objects.
% \newblock {\em ACM Trans. Program. Lang. Syst.}, 12:463--492, July 1990.
% 
% \bibitem{HW90}
% M.~P. Herlihy and J.~M. Wing.
% \newblock Linearizability: a correctness condition for concurrent objects.
% \newblock {\em ACM Trans. Program. Lang. Syst.}, 12:463--492, July 1990.









 


%\vspace{-0.2cm}
\bibitem{HA79}
Hewitt C.E. and Atkinson R.R., 
Specification and Proof Techniques for Serializers. 
{\it IEEE Transactions on Software Engineering}, SE5(1):1-21, 1979. 



%\vspace{-0.2cm}
\bibitem{H74}
Hoare C.A.R., 
Monitors: an Operating System Structuring Concept. 
{\it Communications of the  ACM}, 17(10):549-557, 1974. 




\bibitem{IR08}
Imbs D. and Raynal M.,
A Lock-based STM Protocol  that Satisfies Opacity and Progressiveness. 
{\it 12th Int'l Conference On Principles Of Distributed Systems
(OPODIS'08)},   Springer-Verlag LNCS  \#5401, pp.~226-245, 2008. 


\bibitem{IR09} 
Imbs D. and Raynal M., 
A versatile   STM protocol with Invisible Read Operations
that Satisfies  the  Virtual World Consistency Condition.
{\it  16th  Colloquium   on  Structural   Information   and  Communication
Complexity  (SIROCCO'09)}, Springer Verlag LNCS,  \#5869,  pp. 266-280, 2009. 



% \bibitem{IR09} 
% Imbs, D., Raynal, M.: 
% A versatile   STM protocol with invisible read operations 
% that satisfies  the  virtual world consistency condition.
% In: {\it   16th  Colloquium   on  Structural   Information   and  Communication Complexity  (SIROCCO'09)}, 
% Springer Verlag LNCS   \#5869, pp. 266--280 (2009)



\bibitem{IR09-a}
Imbs D. and Raynal M.,
Provable STM Properties: Leveraging Clock and Locks to Favor Commit 
and Early Abort.
{\it 10th Int'l Conference on Distributed Computing and  Networking
(ICDCN'09)},  Springer-Verlag LNCS  \#5408, pp. 67-78, January 2009. 

% \bibitem{1532612}
% Damien Imbs and Michel Raynal.
% \newblock Provable stm properties: Leveraging clock and locks to favor commit
%   and early abort.
% \newblock In {\em ICDCN '09: Proceedings of the 10th International Conference
%   on Distributed Computing and Networking}, pages 67--78, Berlin, Heidelberg,
%   2009. Springer-Verlag.


% % % % % % % % %
% % % % % % % % %\bibitem{IR08-1}
% % % % % % % % %Imbs D. and Raynal M., 
% % % % % % % % %Help When Needed, but No More: Efficient Read/Write Partial Snapshots.
% % % % % % % % %{\it Proc. 23th Int'l Symposium on  Distributed Computing
% % % % % % % % %(DISC'09)},  Springer-Verlag LNCS   \#5805, pp. 142-156, 2009. 


\bibitem{1612021}
Imbs D. and Raynal M.
\newblock Software transactional memories: An approach for multicore
  programming.
\newblock In {\em PaCT '09: Proceedings of the 10th International Conference on
  Parallel Computing Technologies}, pages 26--40, Berlin, Heidelberg, 2009.
  Springer-Verlag.



\bibitem{abi}
{Intel Corporation}.
\newblock Intel transactional memory compiler and runtime application binary
  interface, May 2009.



\bibitem{sjbc000}
Jones S.~P.
\newblock {\em Beautiful Concurrency}.
\newblock O'Reilly Media, Inc., 2007.





\bibitem{1584013}
% Idit Keidar and Dmitri Perelman.
Keidar I. and Perelman D.
\newblock On avoiding spare aborts in transactional memory.
\newblock In {\em SPAA '09: Proceedings of the twenty-first annual symposium on
  Parallelism in algorithms and architectures}, pages 59--68, New York, NY,
  USA, 2009. ACM.


\bibitem{Kes83}
Kessels J.~L.~W..
\newblock On-the-fly optimization of data structures.
\newblock {\em Comm. ACM}, 26:895--901, 1983.


% \bibitem{Kes83}
% J.~L.~W. Kessels.
% \newblock On-the-fly optimization of data structures.
% \newblock {\em Comm. ACM}, 26:895--901, 1983.




\bibitem{KSF10}
Korland G., Shavit N., and Felber P.
\newblock Deuce: Noninvasive software transactional memory.
\newblock {\em Transactions on HiPEAC}, 5(2), 2010.





\bibitem{L78}
Lamport L., 
Time, Clocks and the Ordering of Events in a Distributed System.
{\em Communications  of the ACM}, 21(7):558-565, 1978.

% % % % % % % % % % %\bibitem{L86}
% % % % % % % % % % %Lamport L., 
% % % % % % % % % % %On interprocess communication. Part 1: Models, Part 2: Algorithms. 
% % % % % % % % % % %{\em Distributed Computing}, 1(2):77-101, 1986.


%\vspace{-0.2cm}
\bibitem{LK08}
Larus J. and Kozyrakis Ch., 
Transactional Memory: Is TM the Answer for Improving Parallel Programming?
{\it Communications of the ACM}, 51(7):80-89, 2008. 

% \bibitem{1364800}
% James Larus and Christos Kozyrakis.
% \newblock Transactional memory.
% \newblock {\em Commun. ACM}, 51(7):80--88, 2008.

\bibitem{Lea}
Lea D.
\newblock Jsr-166 specification request group.


\bibitem{lev:anatomy:transact:2009}
% Yossi Lev, Victor Luchangco, Virendra Marathe, Mark Moir, Dan Nussbaum, and
%   Marek Olszewski.
Lev Y., Luchangco V., MaratheV., Moir M., Nussbaum D. and
  Olszewski M.
\newblock Anatomy of a scalable software transactional memory.
\newblock In {\em TRANSACT~'09: 4th Workshop on Transactional Computing}, feb
  2009.



\bibitem{ma07}
 Maessen, J.-W. and Arvind, M.:
 Store Atomicity for Transactional Memory. 
{\it Electronic  Notes  on Theoretical  Computer Science}, 
174(9):117--137 (2007).



%\vspace{-0.2cm}
\bibitem{MMFLMR11}
Maldonado W.,  Marlier P., Felber P.,  Lawall J., Muller G. and Revi\`ere E., 
Deadline-Aware Scheduling for Software Transactional Memory. 
{\it 41th  IEEE/IFIP Int'l  Conference on  Dependable  Systems and Networks
'DSN'11)}, IEEE CPS Press, June 2011. 


\bibitem{1693465}
% Walther Maldonado, Patrick Marlier, Pascal Felber, Adi Suissa, Danny Hendler,
%   Alexandra Fedorova, Julia~L. Lawall, and Gilles Muller.
Maldonado W., Marlier P., Felber P., Suissa A., Hendler D.,
  Fedorova A., Lawall J.~L. and Muller G.
\newblock Scheduling support for transactional memory contention management.
\newblock In {\em PPoPP '10: Proceedings of the 15th ACM SIGPLAN symposium on
  Principles and practice of parallel programming}, pages 79--90, New York, NY,
  USA, 2010. ACM.



\bibitem{UR84}
% Udi Manbar and Richard~E. Ladner.
Manbar U. and Ladner R.~E.
\newblock Concurrency control in a dynamic search structure.
\newblock {\em ACM Trans. Database Syst.}, 9(3):439--455, 1984.

% \bibitem{UR84}
% U.~Manbar and R.~E. Ladner.
% \newblock Concurrency control in a dynamic search structure.
% \newblock {\em ACM Trans. Database Syst.}, 9(3):439--455, 1984.



\bibitem{Marathe05adaptivesoftware}
% Virendra~J. Marathe, William N.~Scherer Iii, and Michael~L. Scott.
Marathe V.~J., Iii W.N.~S. and  Scott M.~L.
\newblock Adaptive software transactional memory.
\newblock In {\em In Proc. of the 19th Intl. Symp. on Distributed Computing},
  pages 354--368, 2005.



\bibitem{MSHAESS06}
Marathe  V.J.,  Spear M.F., Heriot  C., Acharya A., Eisentatt  D., Scherer
III W.N. and Scott M.L.,
Lowering the Overhead of Software Transactional Memory. 
{\it  Proc 1rt  ACM SIGPLAN  Workshop on  Languages, Compilers  and Hardware
Support for Transactional Computing (TRANSACT'06)}, 2006. 



% \bibitem{Marathe06loweringthe}
% Virendra~J. Marathe, Michael~F. Spear, Christopher Heriot, Athul Acharya, David
%   Eisenstat, William N.~Scherer Iii, and Michael~L. Scott.
% \newblock Lowering the overhead of nonblocking software transactional memory.
% \newblock In {\em Dept. of Computer Science, Univ. of Rochester}, 2006.


\bibitem{blundell06}
 Martin, M.,  Blundell, C. and Lewis, E.:
 Subtleties of Transactional Memory Atomicity Semantics. 
{\it IEEE Computer Architecture  Letters},  5(2):  (2006)



\bibitem{MRL+10}
Mattson T.~G., Riepen M., Lehnig T., Brett P., Haas W., Kennedy P., Howard J.,
  Vangal S., Borkar N., Ruhl G. and Dighe S.
\newblock The 48-core {SCC} processor: the programmer's view.
\newblock In {\em SC}, pages 1--11, 2010.


\bibitem{MS12}
Matveev, A. and  Shavit, N.:
Towards a Fully Pessimistic STM Model. 
In: {\it Proc. Workshop on transactional memory (TRANSACT'12)} (2012)



\bibitem{Mic02}
Michael M.~M.
\newblock High performance dynamic lock-free hash tables and list-based sets.
\newblock In {\em SPAA}, pages 73--82, 2002.

% \bibitem{Mic02}
% M.~M. Michael.
% \newblock High performance dynamic lock-free hash tables and list-based sets.
% \newblock In {\em SPAA}, pages 73--82, 2002.




%\vspace{-0.2cm}
\bibitem{MS96}
Michael M.M.  and Scott M.L., 
Simple, Fast and Practical Blocking and Non-Blocking Concurrent
Queue Algorithms.
{\it  Proc. 15th  Int'l  ACM Symposium on Principles of  Distributed 
Computing (PODC'96)},  ACM Press, pp. 267-275,  1996. 



\bibitem{MTCM07}
Minh, C., Trautmann, M., Chung, J., McDonald, A., Bronson, N., Casper, J., Kozyrakis, C. and Olukotun, K.:
An effective hybrid transactional memory system with strong isolation guarantees.
In: {\it SIGARCH Comput. Archit. News}, 35(2):69--80 (2007)





\bibitem{Moh90}
Mohan C.
\newblock Commit-{LSN}: a novel and simple method for reducing locking and
  latching in transaction processing systems.
\newblock In {\em Proc. of the 16th Int'l Conference on Very Large Data Bases},
  1990.


% \bibitem{Moh90}
% C.~Mohan.
% \newblock Commit-{LSN}: a novel and simple method for reducing locking and
%   latching in transaction processing systems.
% \newblock In {\em Proc. of the 16th Int'l Conference on Very Large Data Bases},
%   1990.


\bibitem{Mos06}
Moss J.~Eliot~B.
\newblock Open nested transactions: Semantics and support.
\newblock In {\em Workshop on Memory Performance Issues}, 2006.


% \bibitem{_opennested}
% J.~Eliot and B.~Moss.
% \newblock Open nested transactions: Semantics and support.



\bibitem{NMA+07}
% Yang Ni, Vijay Menon, Ali-Reza Abd-Tabatabai, Antony~L. Hosking, Richard~L.
%   Hudson, J.~Eliot~B. Moss, Bratin Saha, and Tatiana Shpeisman.
Ni Y., Menon V., Abd-Tabatabai A.-R., Hosking A.~L.,
  Hudson R~L., Moss J.~E.~B., Saha B. and Shpeisman T.
\newblock Open nesting in software transactional memory.
\newblock In {\em Proc. of the 12th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2007.


% \bibitem{NMA+07}
% Y.~Ni, V.~Menon, A.-R. Abd-Tabatabai, A.~L. Hosking, R.~L. Hudson, J.~E.~B.
%   Moss, B.~Saha, and T.~Shpeisman.
% \newblock Open nesting in software transactional memory.
% \newblock In {\em Proc. of the 12th ACM SIGPLAN Symp. on Principles and
%   Practice of Parallel Programming}, 2007.
% 
% \bibitem{1229442}
% Yang Ni, Vijay~S. Menon, Ali-Reza Adl-Tabatabai, Antony~L. Hosking, Richard~L.
%   Hudson, J.~Eliot~B. Moss, Bratin Saha, and Tatiana Shpeisman.
% \newblock Open nesting in software transactional memory.
% \newblock In {\em PPoPP '07: Proceedings of the 12th ACM SIGPLAN symposium on
%   Principles and practice of parallel programming}, pages 68--78, New York, NY,
%   USA, 2007. ACM.


\bibitem{NS98}
Nurmi O. and Soisalon-Soininen E.
\newblock Uncoupling updating and rebalancing in chromatic binary search trees.
\newblock In {\em Proc. of the 10th ACM Symp. on Principles of Database
  Systems}, 1991.

% \bibitem{NS98}
% O.~Nurmi and E.~Soisalon-Soininen.
% \newblock Uncoupling updating and rebalancing in chromatic binary search trees.
% \newblock In {\em Proc. of the 10th ACM Symp. on Principles of Database
%   Systems}, 1991.



\bibitem{NSW87}
Nurmi O., Soisalon-Soininen E., and Wood D.
\newblock Concurrency control in database structures with relaxed balance.
\newblock In {\em Proc. of the 6th ACM Symp. on Principles of Database
  Systems}, 1987.

% \bibitem{NSW87}
% O.~Nurmi, E.~Soisalon-Soininen, and D.~Wood.
% \newblock Concurrency control in database structures with relaxed balance.
% \newblock In {\em Proc. of the 6th ACM Symp. on Principles of Database
%   Systems}, 1987.
% 
% \bibitem{NSW87}
% O.~Nurmi, E.~Soisalon-Soininen, and D.~Wood.
% \newblock Concurrency control in database structures with relaxed balance.
% \newblock In {\em PODS}, 1987.




\bibitem{PA11}
% Victor Pankratius and Ali-Reza Adl-Tabatabai.
Pankratius V. and Adl-Tabatabai A.-R.
\newblock A study of transactional memory vs. locks in practice.
\newblock In {\em Proc. of the 23rd ACM Symp. on Parallelism in Algorithms and
  Architectures}, 2011.



\bibitem{P79}
Papadimitriou Ch.H., 
The Serializability of Concurrent Updates. 
{\it Journal of the ACM},  26(4):631-653, 1979. 


% \bibitem{P79}
% Papadimitriou, Ch.H.:
% The Serializability of Concurrent Updates. 
% {\it Journal of the ACM},  26(4):631--653 (1979) 



\bibitem{PFK10}
Perelman D., Fan R. and Keidar I., 
On Maintaining Multiple versions in STM. 
{\em  Proc.   29th  annual  ACM    Symposium  on Principles  of Distributed
Computing (PODC'10)}, ACM Press, pp.~16-25, 2010. 



\bibitem{Pug90}
Pugh W.
\newblock Skip lists: a probabilistic alternative to balanced trees.
\newblock {\em Commun. ACM}, 33, June 1990.


% \bibitem{Pug90}
% W.~Pugh.
% \newblock Skip lists: a probabilistic alternative to balanced trees.
% \newblock {\em Commun. ACM}, 33, June 1990.




%\vspace{-0.2cm}
\bibitem{R08}
Raynal  M.,
Synchronization is Coming Back, But Is It the Same?
Keynote Speech. {\it  IEEE 22nd Int'l Conference on Advanced 
Information  Networking and Applications (AINA'08)},  pp. 1-10, 2008. 


\bibitem{RFF06}
Riegel T., Felber P., and Fetzer C.
\newblock A lazy snapshot algorithm with eager validation.
\newblock In {\em DISC}, 2006.


\bibitem{RFF07}
Riegel T., Fetzer C. and Felber P.,
Time-based Transactional Memory with Scalable Time Bases.
{\em  Proc.   19th  annual  ACM   Symposium  on  Parallel   Algorithms  and
Architectures (SPAA'07)}, ACM Press, pp.~221-228, 2007.

% \bibitem{10.1109/TPDS.2010.49}
% Pascal Felber, Christof Fetzer, Patrick Marlier, and Torvald Riegel.
% \newblock Time-based software transactional memory.
% \newblock {\em IEEE Transactions on Parallel and Distributed Systems},
%   99(PrePrints), 2010.


%\vspace{-0.2cm}
\bibitem{RSL78}
Rosenkrantz D.J.,  Stearns R.E. and Lewis Ph.M., 
System Level Concurrency Control for Distributed Database Systems.
{\it  ACM Transactions on  Database Systems},  3(2): 178-198, 1978.



\bibitem{RHW10}
% Christopher~J. Rossbach, Owen~S. Hofmann, and Emmett Witchel.
Rossbach C.~J., Hofmann O.~S. and Witchel E.
\newblock Is transactional programming actually easier?
\newblock In {\em Proc. of the 15th ACM SIGPLAN Symp. on Principles and
  Practice of Parallel Programming}, 2010.


\bibitem{1123001}
% Bratin Saha, Ali-Reza Adl-Tabatabai, Richard~L. Hudson, Chi~Cao Minh, and
%   Benjamin Hertzberg.
Saha B., Adl-Tabatabai A.-R., Hudson R.~L., Minh C.~C., and
  Hertzberg B.
\newblock Mcrt-stm: a high performance software transactional memory system for
  a multi-core runtime.
\newblock In {\em PPoPP '06: Proceedings of the eleventh ACM SIGPLAN symposium
  on Principles and practice of parallel programming}, pages 187--197, New
  York, NY, USA, 2006. ACM.




\bibitem{SMSA08}
 Schneider, F., Menon, V., Shpeisman, T. and Adl-Tabatabai, A.:
 Dynamic optimization for efficient strong atomicity.
In: {\it ACM  SIGPLAN Noticers}, 43(10):181--194  (2008)

\bibitem{1073861}
% William~N. Scherer, III and Michael~L. Scott.
Scherer~III W.~N. and Scott M.~L.
\newblock Advanced contention management for dynamic software transactional
  memory.
\newblock In {\em PODC '05: Proceedings of the twenty-fourth annual ACM
  symposium on Principles of distributed computing}, pages 240--248, New York,
  NY, USA, 2005. ACM.




\bibitem{scott:semantics:transact:2006}
% Michael~L. Scott.
Scott M.~L.
\newblock Sequential specification of transactional memory semantics.
\newblock In {\em ACM SIGPLAN Workshop on Transactional Computing}. Jun 2006.
\newblock Held in conjunction with PLDI 2006.



\bibitem{scott07}
 Scott, M.L.,  Spear, M.F., Dalessandro, L. and   Marathe, V.J.:
\newblock Delaunay Triangulation with Transactions and Barriers. 
\newblock In {\it Proc.  10th IEEE Int'l Symposium on Workload Characterization (IISWC '07)},
 IEEE Computer Society, pp. 107--113 (2007)



\bibitem{shpeis07}
 Shpeisman, T.,  Menon, V.,  Adl-Tabatabai, A.R.,  Balensiefer, S.,  Grossman, D.,
 Hudson, R.L., Moore, K.F. and Saha, B.:
\newblock Enforcing isolation and ordering in STM. 
\newblock In {\it ACM  SIGPLAN Noticers}, 42(6):78--88  (2007)

\bibitem{1504199}
% Michael~F. Spear, Luke Dalessandro, Virendra~J. Marathe, and Michael~L. Scott.
Spear M.~F., Dalessandro L., Marathe V.~J. and  Scott M.~L.
\newblock A comprehensive strategy for contention management in software
  transactional memory.
\newblock In {\em PPoPP '09: Proceedings of the 14th ACM SIGPLAN symposium on
  Principles and practice of parallel programming}, pages 141--150, New York,
  NY, USA, 2009. ACM.


\bibitem{spear08}
Spear M.F.,  Dalessandro L.,  Marathe V.J. and  Scott, M.L.
\newblock Ordering-Based Semantics for Software Transactional Memory. 
\newblock In {\it Proc  12th Int'l Conf. on Principles of Distributed Systems 
(OPODIS '08)},  Springer-Verlag LNCS \#5401, pp. 275--294 (2008) 

% \bibitem{spear07}
% Spear, M.F.,  Marathe, V.J.,  Dalessandro, L. and  Scott, M.L.: 
% Privatization techniques for software transactional memory. 
% In: {\it Proc. 26th  annual ACM symposium on Principles of Distributed Computing 
% (PODC '07)}, . ACM press, pp.  338--339, (2007)


\bibitem{spear:privitization:podc:2007}
% Michael~F. Spear, Virendra~J. Marathe, Luke Dalessandro, and Michael~L. Scott.
Spear M.~F., Marathe V.~J., Dalessandro L. and Scott M.~L.
\newblock Privatization techniques for software transactional memory.
\newblock In {\em Proceedings of the 26th PODC ACM Symposium on Principles of
  Distributed Computing}. Aug 2007.








\bibitem{Sha2011}
% Nir Shavit.
Shavit N.
\newblock Data structures in the multicore age.
\newblock {\em Commun. ACM}, 54(3):76--84, 2011.

% \bibitem{Sha11}
% N.~Shavit.
% \newblock Data structures in the multicore age.
% \newblock {\em Commun. ACM}, 54(3):76--84, 2011.
% 
% \bibitem{Sha11}
% N.~Shavit.
% \newblock Data structures in the multicore age.
% \newblock {\em Commun. ACM}, 54(3):76--84, 2011.



\bibitem{ST97}
Shavit N. and Touitou D., 
\newblock Software Transactional Memory. 
\newblock {\it Distributed  Computing}, 10(2):99-116, 1997. 


% \bibitem{224987}
% Nir Shavit and Dan Touitou.
% \newblock Software transactional memory.
% \newblock In {\em PODC '95: Proceedings of the fourteenth annual ACM symposium
%   on Principles of distributed computing}, pages 204--213, New York, NY, USA,
%   1995. ACM.
% 
% 
% \bibitem{shavit95}
%  Shavit, N., Touitou, D.: Software transactional memory. 
% Software Transactional Memory. 
% In: {\it Distributed  Computing}, 10(2):99--116 (1997) 
% 
% 
% %\vspace{-0.2cm}
% \bibitem{ST97}
% Shavit N. and Touitou D., 
% Software Transactional Memory. 
% {\it Distributed  Computing}, 10(2):99-116, 1997. 
% 
% \bibitem{ST95}
% Nir Shavit and Dan Touitou.
% \newblock Software transactional memory.
% \newblock In {\em Proc. of the 14th ACM Symp. on Principles of Distributed
%   Computing}, 1995.



\bibitem{SM93}
Schwarz R. and Mattern F., 
\newblock Detecting  Causal Relationship in Distributed  Computations: in  Search of
the Holy Grail. 
\newblock {\it Distributed Computing}, 7:149-174, 1993. 



%\vspace{-0.2cm}
\bibitem{SSDMS08}
Spear M.F., Silverman M., Dalessandro L., Michael M.M.  and Scott M.L., 
Implementing and Exploiting Inevitability in Software Transactional Memory.
{\it Proc. 37th Int'l Conference on Parallel Processing (ICPP'08)}, 
IEEE Press, 2008. 


\bibitem{ST04}
Sundell H. and Tsigas P.
\newblock Scalable and lock-free concurrent dictionaries.
\newblock In {\em SAC}, pages 1438--1445, 2004.



\bibitem{Sut08}
Sutter H..
\newblock Choose concurrency-friendly data structures.
\newblock {\em Dr. Dobb's Journal}, June 2008.


\bibitem{Tau09}
Taubenfeld G.
\newblock Contention-sensitive data structures and algorithms.
\newblock In {\em DISC}, pages 157--171, 2009.




\bibitem{Val96}
Valois J.~D.
\newblock {\em Lock-free data structures}.
\newblock PhD thesis, Rensselaer Polytechnic Institute, 1996.



%\vspace{-0.2cm}
\bibitem{WF10}
Wamhoff J.-T. and   Fetzer Ch., 
The Universal Transactional Memory Construction. 
{\it Tech Report}, 12 pages, University of Dresden (Germany), 2010. 
 

%\vspace{-0.2cm}
\bibitem{WRFF10}
Wamhoff J.-T.,   Riegel T.,  Fetzer Ch. and  Felber F., 
RobuSTM: A Robust Software Transactional Memory.
{\it Proc. 12th Int'l Symposium on  Stabilization, Safety, and Security 
of Distributed Systems (SSS'10)}, 
Springer Verlag LNCS \#6366, pp.  388-404,  2010. 

%\vspace{-0.2cm}
\bibitem{WSA08}
Welc A., Saha B. and Adl-Tabatabai A.-R., 
Irrevocable Transactions and their Applications. 
{\it  Proc. 20th  Int'l  ACM Symposium on Parallelism in Algorithms 
and Architectures   (SPAA'08)},  ACM Press, pp. 285-296,  2008. 


\bibitem{WGH+07}
Wentzlaff D., Griffin P., Hoffmann H., Bao L., Edwards B., Ramey C.,
  Mattina M., Miao C.-C., Brown J. and Agarwal A.
\newblock On-chip interconnection architecture of the tile processor.
\newblock {\em IEEE Micro}, 27(5):15--31, 2007.


\bibitem{YNW+08}
% Richard~M. Yoo, Yang Ni, Adam Welc, Bratin Saha, Ali-Reza Adl-Tabatabai, and
%   Hsien-Hsin~S. Lee.
Yoo R.~M., Ni Y., Welc A., Saha B., Adl-Tabatabai A.-R. and
  Lee H.-H.~S..
\newblock Kicking the tires of software transactional memory: why the going
  gets tough.
\newblock In {\em Proc. of the 20th ACM Symp. on Parallelism in Algorithms and
  Architectures}, 2008.




















































































































































































































}
\end{thebibliography}
